Screening Sinkhorn Algorithm for Regularized Optimal Transport

### 2019-10-21 final_NeurIPS/appendix.tex
Modify presentation of Proposition 3
correct typo
update

$\norm{u^\star}_\infty \leq R$ and $\norm{v^\star}_\infty\leq R,$ where 

@@

R = \frac{\norm{C}_\infty}{\eta} + \log\big(\frac{n\vee m}{c^2_{\mu\nu}}\big).

@@

&\leq 2\big(R + \norm{u^{\text{sc}}}_\infty +  \norm{v^{\text{sc}}}_\infty + \log\big(\frac{(n\vee m)\varepsilon^2}{c_{\mu\nu}})\big) \big(\norm{\mu^{\text{sc}} - \mu}_1+ \norm{\nu^{\text{sc}} - \nu}_1\big)\nonumber\\

@@

R + \norm{u^{\text{sc}}}_\infty  + \norm{v^{\text{sc}}}_\infty + \log\big(\frac{(n\vee m)\varepsilon^2}{c_{\mu\nu}})\big) = \bigO\Big(\frac{\norm{C}_\infty}{\eta} + \log\Big(\frac{(n\vee m)^2}{nmc_{\mu\nu}^{7/2}}\Big)\Big).

@@

& = \bigO\Big(\Big(\frac{\norm{C}_\infty}{\eta} + \log\Big(\frac{(n\vee m)^2}{nmc_{\mu\nu}^{7/2}}\Big)\Big)(\norm{\mu - {\mu}^{\text{sc}}}_1 + \norm{\nu - \nu^{\text{sc}}}_1 + \omega_{\kappa}\Big).

@@

Note that $\omega_{\kappa}$ tends to zero as $(n_b,m_b)$ tends to $(n,m)$ (i.e., $\kappa \rightarrow 1)$. To more characterize $\omega_\kappa$, the following 
lemma expresses an upper bound with respect to $\ell_1$-norm of $\mu^{\text{sc}}$and $\nu^{\text{sc}}$. 
\begin{lemma}
\label{lemma_bounds_on_marginals}
Let $(u^{\text{sc}}, v^{\text{sc}})$ be an optimal solution of problem~\eqref{screen-sinkhorn_second_def}.
Then one has 
\begin{align}
\label{l1-norm-mu-sc}
\norm{\mu^{\text{sc}}}_1 = \bigO\Big(\frac{n_b\sqrt{m}}{\sqrt{nK_{\min}c_{\mu\nu}}} + (n-n_b)\Big(\frac{m_b}{\sqrt{nmc_{\mu\nu}}K_{\min}^{3/2}} + \frac{m-m_b}{\sqrt{nm}K_{\min}}\Big)\Big),
\end{align}
and
\begin{align}
\label{l1-norm-nu-sc}
\norm{\nu^{\text{sc}}}_1 = \bigO\Big(\frac{m_b\sqrt{n}}{\sqrt{mK_{\min}c_{\mu\nu}}} + (m-m_b)\Big(\frac{n_b}{\sqrt{nmc_{\mu\nu}}K_{\min}^{3/2}} + \frac{n-n_b}{\sqrt{nm}K_{\min}}\Big)\Big).
\end{align}
\end{lemma}

@@

It is worth to note that standard Sinkhorn's alternating minimization cannot be applied for the constrained screened dual problem~\eqref{screen-sinkhorn_second_def}. This appears more clearly while writing its optimality conditions (see Equations~\eqref{i-th-marginal-nu} and~\eqref{i-th-marginal-mu} ).
We resort to a L-BFGS-B algorithm to solve the constrained convex optimization problem on the screened variables~\eqref{screen-sinkhorn_second_def}, but any other efficient solver (e.g., proximal based method or Newton method) could be used. 
In our implementation, we use a tricky method, called \emph{restricted Sinkhorn} (see Algorithm~\ref{restricted_sinkhorn}) as a warm-start initialized point for the L-BFGS-B.
We further  emphasize that, as for the \textsc{Sinkhorn} algorithm, the \textsc{Screenkhorn} algorithm can be accelerated using a GPU implementation\footnote{\url{https://github.com/nepluno/lbfgsb-gpu}} of the L-BFGS-B algorithm~\cite{fei2014-GPU-L-BFGS-B}.

\LinesNotNumbered
\begin{algorithm}[tbp]
\SetNlSty{textbf}{}{.}
\DontPrintSemicolon
\caption{\textsc{Restricted \textsc{Sinkhorn}}$(K,\mu, \nu, u, v)$}
\label{restricted_sinkhorn}
\nl \textbf{set: } $\bar{f}_u = \varepsilon\kappa\, c(K_{I_{\varepsilon, \kappa}, J^\complement_{\varepsilon, \kappa}}), \bar{f}_v  = \varepsilon\kappa^{-1} \,r(K_{I^\complement_{\varepsilon, \kappa}, J_{\varepsilon, \kappa}});$\\
% \nl $\bar{f}_u = \varepsilon\kappa\, c(K_{I_{\varepsilon, \kappa}, J^\complement_{\varepsilon, \kappa}});$\\
% \nl $\bar{f}_v  = \varepsilon\kappa^{-1} \,r(K_{I^\complement_{\varepsilon, \kappa}, J_{\varepsilon, \kappa}});$\\
\nl \For{$t=1,2,3$}{
	$f^{(t)}_v \gets K^\top_{I_{\varepsilon, \kappa}, J_{\varepsilon, \kappa}} u + \bar{f}_v;$\\
	$v^{(t)} \gets \frac{\nu_{J_{\varepsilon, \kappa}}}{\kappa f^{(t)}_v} ;$\\% / (\kappa K{IJ}_v);$\\
	$f^{(t)}_u \gets K_{I_{\varepsilon, \kappa}, J_{\varepsilon, \kappa}} v + \bar{f}_u;$\\
	$u^{(t)} \gets \frac{\kappa \mu_{I_{\varepsilon, \kappa}}}{f^{(t)}_u};$\\
	$u \gets u^{(t)}$, $v \gets v^{(t)}$;
}
\nl \Return{$(u^{(t)}, v^{(t)})$}
\end{algorithm}

@@

\begin{figure*}[htbp]
	\centering
	%~\hfill~	\includegraphics[width=6.cm]{./figs/da_toy_problem.pdf}~\hfill~\\
	~\hfill~\includegraphics[width=6.cm]{./figs/wda_accur_toy.pdf}~\hfill~
	\includegraphics[width=6.cm]{./figs/wda_accur_mnist.pdf}~\hfill~\\
	\caption{Accuracy of a $1$-nearest-neighbour after WDA for the (left) toy problem and, (right) MNIST). We note a slight loss of performance for the toy problem, whereas for MNIST, all approaches yield the same performance. 
		\label{fig:wda_gain}}
\end{figure*}

@@

>>>>

$\norm{u^\star}_\infty \leq A$ and $\norm{v^\star}_\infty\leq A,$ where 

@@

A = \frac{\norm{C}_\infty}{\eta} + \log\big(\frac{n\vee m}{c^2_{\mu\nu}}\big).

@@

&\leq 2\big(A + \norm{u^{\text{sc}}}_\infty +  \norm{v^{\text{sc}}}_\infty + \log\big(\frac{(n\vee m)\varepsilon^2}{c_{\mu\nu}})\big) \big(\norm{\mu^{\text{sc}} - \mu}_1+ \norm{\nu^{\text{sc}} - \nu}_1\big)\nonumber\\

@@

A + \norm{u^{\text{sc}}}_\infty  + \norm{v^{\text{sc}}}_\infty + \log\big(\frac{(n\vee m)\varepsilon^2}{c_{\mu\nu}})\big) = \bigO\Big(\frac{\norm{C}_\infty}{\eta} + \log\Big(\frac{(n\vee m)^2}{nmc_{\mu\nu}^{7/2}}\Big)\Big).

@@

& = \bigO\big(R(\norm{\mu - {\mu}^{\text{sc}}}_1 + \norm{\nu - \nu^{\text{sc}}}_1 + \omega_{\kappa})\big).

@@

Note that $\omega_{\kappa}$ tends to zero as $(n_b,m_b)$ tends to $(n,m)$ (i.e., $\kappa \rightarrow 1)$. To more characterize $\omega_\kappa$, the following 
lemma expresses an upper bound with respect to $\ell_1$-norm of $\mu^{\text{sc}}$and $\nu^{\text{sc}}$.

\begin{lemma}
\label{lemma_bounds_on_marginals}
Let $(u^{\text{sc}}, v^{\text{sc}})$ be an optimal solution of problem~\eqref{screen-sinkhorn_second_def}.
Then one has 
\begin{align}
\label{l1-norm-mu-sc}
\norm{\mu^{\text{sc}}}_1 = \bigO\Big(\frac{n_b\sqrt{m}}{\sqrt{nK_{\min}c_{\mu\nu}}} + (n-n_b)\Big(\frac{m_b}{\sqrt{nmc_{\mu\nu}}K_{\min}^{3/2}} + \frac{m-m_b}{\sqrt{nm}K_{\min}}\Big)\Big),
\end{align}
and
\begin{align}
\label{l1-norm-nu-sc}
\norm{\nu^{\text{sc}}}_1 = \bigO\Big(\frac{m_b\sqrt{n}}{\sqrt{mK_{\min}c_{\mu\nu}}} + (m-m_b)\Big(\frac{n_b}{\sqrt{nmc_{\mu\nu}}K_{\min}^{3/2}} + \frac{n-n_b}{\sqrt{nm}K_{\min}}\Big)\Big).
\end{align}
\end{lemma}

@@

It is worth to note that standard Sinkhorn's alternating minimization cannot be applied for the constrained screened dual problem~\eqref{screen-sinkhorn_second_def}. This appears more clearly while writing its optimality conditions (see Equations~\eqref{i-th-marginal-nu} and~\eqref{i-th-marginal-mu} ).
We resort to a L-BFGS-B algorithm to solve the constrained convex optimization problem on the screened variables~\eqref{screen-sinkhorn_second_def}, but any other efficient solver (e.g., proximal based method or Newton method) could be used. 
In our implementation, we choose a clever starting point of the L-BFGS-B algorithm based on a method, called \emph{restricted Sinkhorn} (see Algorithm~\ref{restricted_sinkhorn}), which is a Sinkhorn algorithm applied on
the active dual variables. While simple and efficient the solution of this 
\emph{restricted Sinkhorn} algorithm does not satisfy the lower bound constraints of Problem \eqref{screen-sinkhorn_second_def}.

We further  not that, as for the \textsc{Sinkhorn} algorithm, our \textsc{Screenkhorn} algorithm can be accelerated using a GPU implementation\footnote{\url{https://github.com/nepluno/lbfgsb-gpu}} of the L-BFGS-B algorithm~\cite{fei2014-GPU-L-BFGS-B}.

\LinesNotNumbered
\begin{algorithm}[tbp]
\SetNlSty{textbf}{}{.}
\DontPrintSemicolon
\caption{\textsc{Restricted \textsc{Sinkhorn}}$(K,\mu, \nu, u, v)$}
\label{restricted_sinkhorn}
\nl \textbf{set:} $\bar{f}_u = \varepsilon\kappa\, c(K_{I_{\varepsilon, \kappa}, J^\complement_{\varepsilon, \kappa}}), \bar{f}_v  = \varepsilon\kappa^{-1} \,r(K_{I^\complement_{\varepsilon, \kappa}, J_{\varepsilon, \kappa}});$\\
% \nl $\bar{f}_u = \varepsilon\kappa\, c(K_{I_{\varepsilon, \kappa}, J^\complement_{\varepsilon, \kappa}});$\\
% \nl $\bar{f}_v  = \varepsilon\kappa^{-1} \,r(K_{I^\complement_{\varepsilon, \kappa}, J_{\varepsilon, \kappa}});$\\
\nl \For{$t=1,2,3$}{
	$f^{(t)}_v \gets K^\top_{I_{\varepsilon, \kappa}, J_{\varepsilon, \kappa}} u + \bar{f}_v;$\\
	$v^{(t)} \gets \frac{\nu_{J_{\varepsilon, \kappa}}}{\kappa f^{(t)}_v} ;$\\% / (\kappa K{IJ}_v);$\\
	$f^{(t)}_u \gets K_{I_{\varepsilon, \kappa}, J_{\varepsilon, \kappa}} v + \bar{f}_u;$\\
	$u^{(t)} \gets \frac{\kappa \mu_{I_{\varepsilon, \kappa}}}{f^{(t)}_u};$\\
	$u \gets u^{(t)}$, $v \gets v^{(t)}$;
}
\nl \Return{$(u^{(t)}, v^{(t)})$}
\end{algorithm}

@@

\begin{figure*}[htbp]
	\centering
	%~\hfill~	\includegraphics[width=6.cm]{./figs/da_toy_problem.pdf}~\hfill~\\
	~\hfill~\includegraphics[width=6.5cm]{./figs/wda_accur_toy.pdf}~\hfill~
	\includegraphics[width=6.5cm]{./figs/wda_accur_mnist.pdf}~\hfill~\\
	\caption{Accuracy of a $1$-nearest-neighbour after WDA for the (left) toy problem and, (right) MNIST). We note a slight loss of performance for the toy problem, whereas for MNIST, all approaches yield the same performance. 
		\label{fig:wda_gain}}
\end{figure*}

\begin{figure*}[htbp]

\begin{center}

	~\hfill~\includegraphics[width=6.6cm]{./figs/wda_accur_toy_eta01_K1.pdf}~\hfill~
	\includegraphics[width=6.5cm]{./figs/wda_gain_toy_eta01_K1.pdf}~\hfill~\\

	~\hfill~\includegraphics[width=6.5cm]{./figs/wda_accur_toy_eta01_K5.pdf}~\hfill~
	\includegraphics[width=6.5cm]{./figs/wda_gain_toy_eta01_K5.pdf}~\hfill~\\

	\caption{(top-left) Accuracy and (bottom-right) computational time gain on the toy dataset for $\eta=0.1$ and $1$-nearest-neighbour. (bottom) accuracy and gain but for a $5$-nearest-neighbour. We can note that a  slight loss of performances occur for larger training set sizes especially for $1$-nearest-neighbour. Computational gains increase with the dataset size and
		is on average of the order of magnitude. \ar{Checker 3500 K5(plus d'iter)} 
		\label{fig:wda_gain_toy_eta01}}

\end{center}

\end{figure*}

@@

\begin{figure*}[htbp]
	\centering

	\includegraphics[width=6.5cm]{./figs/da_regpath_accur_mnist_reg1.pdf}
	\includegraphics[width=6.5cm]{./figs/da_regpath_time_mnist_reg1.pdf}

	\caption{OT Domain adaptation MNIST to USPS : (left) Accuracy and (right) running time of \textsc{Sinkhorn} and \textsc{Screenkhorn} for the best performing (on average of $10$ trials) hyperparameter $\ell_{p,1}$ chosen among the set $\{0.1, 1,5,10 \}$. We can note that in this situation, there is not loss of accuracy while our \textsc{Screenkhorn} is still about an order of magnitude more efficient than Sinkhorn.}
\end{figure*}

### 2019-10-21 final_NeurIPS/section_5_numeric_expes.tex
update

	\caption{Wasserstein Discriminant Analysis : running time gain for a toy dataset and for MNIST as a function of the number of examples and the data decimation factor in \textsc{Screenkhorn}.}
	\label{fig:wda}
\end{figure*}
\begin{figure*}[t]
	\centering

	% --- \includegraphics[width=0.43\textwidth]{./figs/da_gain_mnist_regcl1.pdf}\hspace{1cm}
	% --- \includegraphics[width=0.43\textwidth]{./figs/da_gain_mnist_regcl10.pdf}
	\includegraphics[width=0.35\textwidth]{./figs/da_gain_mnist_regcl1.pdf}\hspace{1cm}
	\includegraphics[width=0.35\textwidth]{./figs/da_gain_mnist_regcl10.pdf}
	\caption{OT Domain adaptation : running time gain  for MNIST as a function of the number of examples and the data decimation factor in \textsc{Screenkhorn}. Hyperparameter values (left) $1$. (right) $10$.}
	\label{fig:otda}
\end{figure*}

>>>>

	\caption{Wasserstein Discriminant Analysis : running time gain for (left) a toy dataset and (right) MNIST as a function of the number of examples and the data decimation factor in \textsc{Screenkhorn}.}
	\label{fig:wda}
\end{figure*}
\begin{figure*}[t]
	\centering

	% --- \includegraphics[width=0.43\textwidth]{./figs/da_gain_mnist_regcl1.pdf}\hspace{1cm}
	% --- \includegraphics[width=0.43\textwidth]{./figs/da_gain_mnist_regcl10.pdf}
	\includegraphics[width=0.35\textwidth]{./figs/da_gain_mnist_regcl1.pdf}\hspace{1cm}
	\includegraphics[width=0.35\textwidth]{./figs/da_gain_mnist_regcl10.pdf}
	\caption{OT Domain adaptation : running time gain  for MNIST as a function of the number of examples and the data decimation factor in \textsc{Screenkhorn}. Group-lasso hyperparameter values (left) $1$. (right) $10$.}
	\label{fig:otda}
\end{figure*}
